---
title: "APSTA Week 02 Exercise Solutions"
output: pdf_document
---
**1.** The data set `pi2000` (UsingR) contains the first 2,000 digits of $\pi$. What is the percentage of digits that are 3 or less? What is the percentage of the digits are 5 or more?  
*Solution*:  
```{r, message=FALSE}
library(UsingR)
three_leq <- length(pi2000[pi2000 <= 3])
five_geq <- length(pi2000[pi2000 >= 5])
sprintf(
  "Around %s%% of the digits are three or less, while around %s%% are five or more.",
  round(
    (three_leq/2000)*100, 3
  ), 
  round(
    (five_geq/2000)*100, 3
  )
)
```

**2.** Load the `starwars` dataset from the `dplyr` package. 
The dataset has 87 characters from the StarWars universe with 13 features.  **Note** there are missing data in the database represented by `NA` values.  
```{r}
starwars <- dplyr::starwars
```

a) What is the homeworld of Mace Windu?  
*Solution*:  
```{r}
sprintf("Mase Windu is from %s", starwars[starwars$name == "Mace Windu",]$homeworld)
```

b) How many droids are in the dataset?  
*Solution*:  
```{r}
sprintf(
  "The dataset includes %s different droids.", 
  nrow(starwars[starwars$species == "Droid",])
)
```

c) Who are the shortest and tallest humans in dataset?  
*Solution*:  
```{r}
min_height <- min(starwars$height, na.rm=TRUE)
max_height <- max(starwars$height, na.rm=TRUE)
shortest <- starwars[starwars$height == min_height,][1,]$name
tallest <- starwars[starwars$height == max_height & !is.na(starwars$height),]$name
sprintf(
  "Smallest person in Star Wars: %s. Tallest person in Star Wars: %s.", 
  shortest, tallest
)
```

d) What is the mean and standard deviation of the height all humans in the starwars database?  
```{r}
mean_height <- mean(starwars[starwars$species == "Human",]$height, na.rm=TRUE)
SD_height <- sd(starwars[starwars$species == "Human",]$height, na.rm=TRUE)
sprintf("The mean height of Humans in Star Wars is %scm with a standard deviation of %s.",
        round(mean_height, 1), round(SD_height, 1))
```

**3.** The following table shows the results of a survey of 10 pirates. 
In addition to some basic demographic information, the survey asked each pirate “What is your favorite superhero?”" and “How many tattoos do you have?”"

| Name|  	Sex |	Age |	Superhero |	Tattoos|
|---------|---|----| --------- | ---- |
|Astrid 	| F |	30 |	Batman 	 |  11  |
|Lea 	    | F |	25 |	Superman |	15  |
|Sarina 	| F |	25 |	Batman 	 |  12  |
|Remon 	  | M |	29 |	Spiderman| 	5   |
|Letizia 	| F |	22 |	Batman 	 |  65  |
|Babice  	| F |	22 |	Antman 	 |  3   |
|Jonas 	  | M |	35 |	Batman 	 |  9   |
|Wendy 	  | F |	19 |	Superman |	13  |
|Niveditha| F |	32 |	Maggott  |	900 |
|Gioia 	  | F |	21 |	Superman |	0   |

a) Combine the data into a single dataframe.  
Complete all the following exercises from the dataframe!  
*Solution*:  
```{r}
pirates <- data.frame(
  name = c("Astrid", "Lea", "Sarina", "Remon", "Letizia", "Babice",
           "Jonas", "Wendy", "Niveditha", "Gioia"),
  sex = c("F", "F", "F", "M", "F", "F", "M", "F", "F", "F"),
  age = c(30, 25, 25, 29, 22, 22, 35, 19, 32, 21),
  superhero = c("Batman", "Superman", "Batman", "Spiderman", "Batman",
                "Antman", "Batman", "Superman", "Maggott", "Superman"),
  tattoos = c(11, 15, 12, 5, 65, 3, 9, 13, 900, 0)
)
head(pirates)
```

b) What was the mean age of female and male pirates separately?  
*Solution*:  
```{r}
male_mean <- mean(pirates[pirates$sex == "M",]$age)
female_mean <- mean(pirates[pirates$sex == "F",]$age)
sprintf("Male mean: %s. Female mean: %s.", male_mean, female_mean)
```

c) Add a new column to the dataframe called `tattoos.per.year` which shows how many tattoos each pirate has for each year in their life. Which pirate had the most number of tattoos per year?  
*Solution*:  
```{r}
pirates$tattoos.per.year <- pirates$tattoos / pirates$age
tattoo_freak = pirates[pirates$tattoos.per.year == max(pirates$tattoos.per.year),][1,]
sprintf("%s had %s tattoos by the age of %s. That's a whopping %s tattoos per year!",
        tattoo_freak$name, 
        tattoo_freak$tattoos, 
        tattoo_freak$age, 
        tattoo_freak$tattoos.per.year)
```

d) What was the median number of tattoos of pirates over the age of 20 whose favorite superhero is Spiderman?  
*Solution*:  
```{r}
filtered <- pirates[
  pirates$age > 20 &
  pirates$superhero == "Spiderman",
]
sprintf(
  "Median tattoos: %s (though there is only one person who fits the criteria).",
  median(filtered$tattoos)
)
```

e) Make a boxplot of the age distribution of the pirates  
*Solution*:  
```{r}
boxplot(
  pirates$age,
  xlab="Age", ylab="Pirates",
  horizontal=TRUE
)
```

f) Make a piechart showing the number of pirates which has each superhero as their favorite.  
*Solution*:  
```{r}
pie(table(pirates$superhero))
```