---
title: "Distributions, expectation, variance - solutions"
author: "Anna Maria Gnat"
date: "2023-05-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Mock 2023 - 2. Continuous distributions, Expectation, Variance (6 pts)

Let $X$ be a continuous random variable with the probability density function

$$
  f_{X}(x)= \begin{cases} \frac{3}{4}(1-x^2), \quad  & x \in [-1,1], \\
  0, \quad & \textrm{otherwise}.
  \end{cases}
$$

(a) (2pts) Compute the expected value of $X$.

(b) (2pts) Compute the variance of $X$.

(c) (2pts) Compute the probability $P(X>0.5)$.

#### Solution

(a) 

```{r}
# function to integrate
integrand<-function(x){x*(3/4)*(1-x^2)}

#integrate the function
integral <- integrate(integrand,lower=-1,upper=1)
integral
```

General formula: $E[X] = \int_{-\infty}^{\infty}xf_X(x) dx$\
We are only interested in range [-1, 1] as the function is 0 otherwise:\
$$E[X] = \int_{-1}^{1}x * \frac{3}{4}(1-x^2) dx = \frac{3}{4}\int_{-1}^1 (x - x^3) dx = \frac{3}{4}[\frac{x^2}{2}-\frac{x^4}{4}]_{-1}^1 = 0$$

(b) 

```{r}
# function to integrate
integrand<-function(x){x^2*(3/4)*(1-x^2)}

#integrate the function
integral <- integrate(integrand,lower=-1,upper=1)
integral
```

General formula: ${Var}[X] = E[X^2]-(E[X])^2$\
$E[X^2] = \int_{-1}^{1}x^2 * \frac{3}{4}(1-x^2) dx =0.2$ ${Var}[X] = 0.2 -(0)^2 = 0.2$

(c) 

```{r}
# function to integrate
integrand<-function(x){(3/4)*(1-x^2)}

#integrate the function
integral <- integrate(integrand,lower=0.5,upper=1)
integral
```

$P(X>0.5) = \int_{0.5}^{1}x * \frac{3}{4}(1-x^2) dx = 0.15625$

### Mock 2022 - 2. Expectation, Variance, Discrete Distributions (6 pts)

Let us consider the experiment of independently throwing two fair dice characterised by the discrete random variables $X$ and $Y$, respectively. The discrete random variables $X$ and $Y$ take the values $a=1,2,3,\ldots,6$ and $b=1,2,3,\ldots,6$, respectively.

(a) Compute the expected value for the product $Z=XY$. (2pts)

(b) Write down the probability mass function for the the discrete random variable $Z$ defined by the product $Z=XY$. (2pts)

(c) Compute the variance of the product $Z=XY$. (2pts)

#### Solution

(a) 

$Z=XY$\
Since X and Y are independent, the expectation of Z is: $E[Z] = E[X] * E[Y]$ $E[X] = \sum_{i} a_i* p(a_i) = 1 * \frac{1}{6} + 2 * \frac{1}{6} + 3 * \frac{1}{6} + 4 * \frac{1}{6} + 5 * \frac{1}{6} + 6 * \frac{1}{6} = \frac{21}{6} = \frac{7}{2}$\
$E[Z] = \frac{7}{2} * \frac{7}{2} = \frac{49}{4}$

(b) 

Possibles outcome of X\*Y are:

```{r}
outcomes <- outer(c(1,2,3,4,5,6), c(1,2,3,4,5,6), "*")
outcomes
x <- table(outcomes)
x
```

Therefore the PMF for Z is: $$
p(c) =P(Z=c) = 
\begin{cases} 
\frac{1}{36} & \text{when } c=1,9,16,25,36\\
\frac{1}{18} & \text{when } c=2,3,5,8,10,15,18,20,24,30\\
\frac{1}{12} & \text{when } c=4\\
\frac{1}{9} & \text{when } c=6,12\\
0 & \text{otherwise.}
\end{cases}.
$$ (c) Formula for variance: $Var(Z) = E[Z^2] - (E[Z])^2$\
We need to calculate $E[Z^2]$ which $= E[X^2] * E[Y^2]$ and $(E[Z])^2 = (E[XY])^2$\
$E[X^2]= E[Y^2] = \frac{1}{6}*(1^2 + 2^2 + 3^2 + 4^2 + 5^2 + 6^2) =* \frac{1}{6}(91) = \frac{91}{6}$\
$E[Z^2] = E[X^2]* E[Y^2] = (\frac{91}{6})^2 = \frac{8281}{36}$\
From task b)\
$(E[Z])^2 = (E[XY])^2 = (\frac{49}{4})^2 = \frac{2401}{16}$\
Now, the $Var(Z)$ is:\
$Var(Z) = \frac{8281}{36} - \frac{2401}{16} \approx 79.965$

### Mock 2021 - 2. Expectation, Joint Distributions, Independence (5 pts)

Let (X, Y ) be a pair of continuous random variables with the joint density function:

$$
  f_{}(x,y)= \begin{cases} 1  \quad & if \quad (x,y) \in [0,1] \times [1,2] \\
  0, \quad & \textrm{otherwise}.
  \end{cases}
$$

(a) Compute the expected value of the vector (X, Y ). (1pt)

(b) Compute the covariance between the the random variables X and Y. AreX and Y correlated? (2pts)

(c) Find the the marginal densities of the random variables X and Y. Are X and Y are independent? (2pts)

### Re-exam 2020 - 2.Continuous Random Variable (6 pts)

Assume that the score a student can get in an exam is a real number in the interval [0,100], and the student passes the exam if the score is at least 55. Assume that we model the score of the exam by the random variable X that has the probability density function Certainly! Here's the formula written in LaTeX: $$
f(x) = \begin{cases}
x - \frac{{4p}}{{5}}, & \text{for } 80p < x \leq 80p + 10 \\
-\frac{{x}}{{100}} + \frac{{4p}}{{5}} + \frac{{1}}{{5}}, & \text{for } 80p + 10 < x < 80p + 20 \\
0, & \text{otherwise}
\end{cases}
$$ where p, 0 $\le$ p $\le$ 1 is a parameter that describes how well the student prepared for the exam: p = 0 refers to no preparation and p = 1 excellent preparation. Assume that p = 0.5, i.e., the student did not prepare very well. Using this model, compute

(a) the expected score the student will get, and (3pts)

(b) the probability that the student will pass the exam. (3pts)

### Exam 2020 - 2 Expectation, Joint Distributions, Independence (5 pts)

Let (X, Y ) be a pair of continuous random variables with the joint density function

$$
f(x,y) = \begin{cases} 1 & if (x,y) \in [0,1] x [1,2]\\
0 &otherwise 
\end{cases} 
$$

(a) Compute the expected value of the vector (X, Y ). (1pt)

(b) Compute the covariance between the the random variables X and Y. Are X and Y correlated? (2pts)

<!-- -->

c)  Find the the marginal densities of the random variables X and Y. Are X and Y are independent? (2pts)
