---
title: "APSTA Week 13 Exercises"
output: pdf_document
---
# 1. One sample $t$-test (T)  
We perform a t-test for the null hypothesis $H_0:\mu=10$ by means of a dataset
consisting of $n=16$ elements with sample mean $11$ and sample variance $4$. We
use significance level $0.05$.  
a. Should we reject the null hypothesis in favor of $H_1:\mu\neq10$?  
b. What if we test against $H_1:\mu>10$?  
  
*Solution*:  
The test statistic we use is
$$
T=\frac{\bar{X}_n-\mu_0}{S_n/\sqrt{n}}
$$
where $\bar{x}_n=11$, $\mu_0=10$, $s_n=\sqrt{4}$ and $n=16$.  
This test statistic has a $t(n-1)$ distribution. Since
$$
\mathrm{P}(T\leq-t_{n-1,\alpha/2}\ \ \ \text{or}\ \ \ T\geq t_{n-1,\alpha/2})=\alpha,
$$
we can reject the null hypothesis of $T$ goes outside of these bounds. Here, $\alpha=0.05$.  
$$
\mathrm{P}(T\leq-t_{15,0.025}\ \ \ \text{or}\ \ \ T\geq t_{15,0.025})=0.05.
$$
The critical value can be found using the `qt` function:  
```{r}
qt(1-0.05/2, 15)
```
Now we calculate the realisation of the test statistic.  
$$
t=\frac{\bar{x}_n-\mu_0}{s_n/\sqrt{n}}=\frac{11-10}{\sqrt{4}/\sqrt{16}}=2.
$$  
Since $2\ngeq2.132$, we reject the null hypothesis.  
To check the one-sided probability in the hypothesis $H_1:\mu>10$, we use $t_{n-1,\alpha}$ instead of $t_{n-1,\alpha/2}$.  
```{r}
qt(1-0.05, 15)
```
Since $2\geq1.753$, we cannot reject $H_0$ in favour of $H_1:\mu>10$. We can only reject $H_0$ in favour of $H_1:\mu\neq10$.  
(This makes sense because the two-sided version takes into account the extremes on both sides. By removing the lower extreme, we can decrease the upper extreme under the same confidence.)  
  
# 2. Major League Baseball (R)  
The data set OBP (Using R) contains on-base percentages for the $2002$ Major
League Baseball season. Do a significance test to see wheter the mean on-base
percentage is $0.330$ against a two-sided alternative.  
  
*Solution*:  
```{r}
# Import the dataset
library(UsingR)
summary(OBP)
```
We want to test the alternative hypothesis that the mean $\mu$ is not equal to $\mu_0=0.330$. Therefore, we're testing  
$$
H_0:\mu=0.330\ \ \ \text{against}\ \ \ H_1:\mu\neq0.330
$$
Let's see if the data looks normally distributed:
```{r}
length(OBP)
plot(density(OBP))
shapiro.test(OBP)
```
The size of the dataset is not large enough for a Z-test, and the shapiro test tells us that the data is not very normally distributed. We will therefore bootstrap the studentised mean.  
```{r}
n <- length(OBP)
sqrt_n <- sqrt(n)
x_bar <- mean(OBP)

t_s_samples <- c()
for (i in 1:10000) {
  bootstrap_sample <- sample(OBP, size=n, replace=TRUE)
  
  bootstrap_x_bar <- mean(bootstrap_sample)
  bootstrap_sd <- sd(bootstrap_sample)
  t_s <- (bootstrap_x_bar - x_bar) / (bootstrap_sd / sqrt_n)
  
  t_s_samples <- c(t_s_samples, t_s)
}
```
Now we can find $c_l^*$ and $c_u^*$ such that $\mathrm{P}(T^*\leq c_l^*\ \ \ \text{or}\ \ \ T^*\geq c_u^*)=\alpha$. At a significance of $0.05$, $\alpha=0.05$.  
```{r}
c_s <- quantile(t_s_samples, probs=c(0.05/2, 1-0.05/2))
t <- (x_bar - 0.330) / (sd(OBP) / sqrt_n)

cat("Test statistic realisation:", t)
cat("\nLower/upper bound @ significance 0.05:", c_s)
```
This value is not outside of the critical region, so we do not reject the null hyothesis.  
  
# 3. Easter eggs (T & R)  
Assume that you got six similar Easter eggs with $20g$ of chocolate reported in
each. After taking one more lecture in Applied Statistics, you want to further
investigate whether it is plausible that the eggs really contain $20g$ chocolate or if
the egg producer is cheating. You weight the eggs and obtain the following six
observations for the chocolate weight:  
$$
\begin{aligned}
\text{Chocolate contents (g)} \\
[20.1, 19.1, 18.2, 20.2, 19.6, 19.1]
\end{aligned}
$$  
You may assume that you measurement is a realization of a random sample from
a normal distribution $N(\mu,\sigma^2)$, where $\mu$ represents the true average contents.  
(a) Formulate the appropriate null hypothesis and alterantive hypothesis.  
  
*Solution*:  
The null hypothesis is that the mean is truly equal to $20g$ as they claim, that is, $H_0:\mu=20$. We are interested in whether or not the company is scamming us, so we only care if there is less chocolate than advertised, that is $H_1:\mu<20$.  
  
(b) Which test is appropriate for testing the hypothesis? Explain why.  
  
*Solution*:  
We assume the measurements are from an $N(\mu,\sigma^2)$ distribution, and we only have a single sample, so we use the single-sample t-test.  
  
(c) Compute the value of the test statistic and report your conclusion at significance level $\alpha=0.05$.  
  
*Solution*:  
The test statistic is the studentised mean:  
$$
T=\frac{\bar{X}_n-\mu_0}{S_n/\sqrt{n}}
$$
First we want to know the mean, standard deviation and square root of $n$:
```{r}
data <- c(20.1, 19.1, 18.2, 20.2, 19.6, 19.1)
cat("Mean:", mean(data))
cat("\nStandard deviation:", sd(data))
cat("\nsqrt(n):", sqrt(length(data)))
```
And then we calculate our realisation of the test statistic:  
$$
t=\frac{\bar{x}_n-\mu_0}{s_n/\sqrt{n}}=\frac{19.38\bar{3}-20}{0.747/2.45}=-\frac{0.61\bar{6}}{0.305}=-2.023
$$  
Finally, we need to find the critical value $t_{n-1,\alpha/2}$ where $\alpha=0.05$. This can be done using `qt`:
```{r}
ct <- -qt(1-0.05, 6-1)
cat("Critical value:", ct)
```
As our value of $-2.023$ is less than the critical value $-2.015$, the reject the null hypothesis. The manufacturer is actually putting less than $20g$ of chocolate in the eggs.  
  
(d) Compute the corresponding left ail $p$-value. Is it likely to observe these measurements under the null hypothesis?  
  
*Solution*:  
The p-value is calculated by $\mathrm{P}(T\leq t)$. We can calculate this using the `pt` function:  
```{r}
n <- length(data)
t <- (mean(data)-20) / (sd(data)/sqrt(n))
p <- pt(t, n-1)
cat("p-value:", p)
```
So the p-value is only very slightly below the significance level $0.05$.  
  
# 4. Two-sample $t$-test (T)  
The data in Table $28.3$ (pp. $425$, Dekking et al. ($2010$)) represents salaries (in
pounds Sterling) in $72$ randomly selected advertisements in The Guardian (April
$6$, $1992$). When the range was given in the advertisement, the midpoint of the
range is reproduced in the table. The data are salaries corresponding to two
kinds of occupations ($n=m=72$): ($1$) Creative, media, and marketing and
($2$) education. The sample mean and sample variance of the two datasets are,
respectively:  
(1) $\bar{x}_{72}=17410$ and $s_x^2=41258741$,  
(2) $\bar{y}_{72}=19818$ and $s_y^2=50744521$,  
  
Supposed that the datasets are modeled as realizations of normal distributions
with expectations $\mu_1$ and $\mu_2$, which represent the salaries for occupations ($1$)
and ($2$).  
(a) Test the null hypothesis that the salary for both occupations is the same
at level $\alpha=0.05$ under the assumption of equal variances. Formulate
the proper null and alternative hypotheses, compute the value of the test
statistic, and report your conclusion.  
  
*Solution*:  
We want to test $H_0:\mu_1=\mu_2$ against $H_1:\mu_1\neq=\mu_2$. We assume equal variance and normality, so we perform the two sample t-test with test statistic
$$
T_p=\frac{\bar{X}_n-\bar{Y}_m}{S_p}
$$
where
$$
S_p^2=\frac{(n-1)S^2_X+(m-1)S^2_Y}{n+m-2}\left(\frac{1}{n}+\frac{1}{m}\right).
$$
The test statistic has a $t(n+m-2)$ distribution. To test at level $\alpha=0.05$ (and since $n=5=72$) we need the critical value $t_{142,0.05/2}$.  
```{r}
ct <- qt(1-0.05/2, 142)
cat("Critical value:", ct)
```
The critical region is therefore any value outside of $[-1.976811,1.97681]$. We calculate the test statistic:  
$$
t_p=\frac{\bar{x}_{72}-\bar{y}_{72}}{s_p}=\frac{17410-19818}{\frac{71(41258741)+71(50744521)}{142}\cdot\frac{2}{72}}=-0.00189
$$
This value is outside the critical region, so the null hypothesis is not rejected.  
  
(b) Do the same without the assumption of equal variances.  
  
*Solution*:  
