---
title: "APSTA Week 10 Exercises"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# 1. Unbiased estimators (Theory)

Consider a random sample $X_1,\dots,X_n$ from a uniform distribution in
the interval $-\theta,\theta$, where $\theta$ is an unknwon parameter.
You are interested in estimating the values of $\theta$.\
a. Show that $$
\hat{\Theta}=\frac{2}{n}(|X_1|+|X_2|+\dots+|X_n|)
$$ is an unbiased estimator for $\theta$. *Hint*: you may need to use
the *change of variable formula* (cfr. Chapter 7 of the book).\
b. Consider instead the problem of estimating $\theta^2$. Show that $$
T=\frac{3}{n}(X^2_1+X^2_1+\dots+X^2_n)
$$ is an unbiased estimator for $\theta^2$\
c. Is $\sqrt{T}$ an unbiased estimator for $\theta$? If not, discuss
whether it has positive or negative bias.

*Solution*:\
a.\
To check whether $\hat{\Theta}$ is unbiased, we need to find its
expectation. $$
\begin{aligned}
\mathrm{E}\left[\hat{\Theta}\right]&=\mathrm{E}\left[\frac{2}{n}(|X_1|+|X_2|+\dots+|X_n|)\right] \\
&=\mathrm{E}\left[\frac{2}{n}|X_1|+\frac{2}{n}|X_2|+\dots+\frac{2}{n}|X_n|\right]
\end{aligned}
$$ I use linearity of expectations:\
$$
\begin{aligned}
&=\mathrm{E}\left[\frac{2}{n}|X_1|\right]+\mathrm{E}\left[\frac{2}{n}|X_2|\right]+\dots+\mathrm{E}\left[\frac{2}{n}|X_2|\right] \\
&=\frac{2}{n}(\mathrm{E}[|X_1|]+\mathrm{E}[|X_2|]+\dots+\mathrm{E}[|X_n|])
\end{aligned}
$$ All the $X_i$ are from the same $U(-\theta,\theta)$ distribution. We
want the expectation of $|X_i|$, so we need to use the change of
variable formula. The probability density function $f(x)$ of the uniform
distribution is $f(x)=1/(\beta-\alpha)$ where $\alpha$ is the lower
bound, here $-\theta$, and $\beta$ the upper bound, here $\theta$.\
$$
\begin{aligned}
\mathrm{E}[|X_i|]&=\int_{-\theta}^\theta |x|f(x)\ \mathrm{d}x \\
&=\int_{-\theta}^\theta |x|\frac{1}{\theta-(-\theta)}\ \mathrm{d}x \\
&=\int_{-\theta}^\theta \frac{|x|}{2\theta}\ \mathrm{d}x \\
&=\frac{1}{2\theta}\int_{-\theta}^\theta |x|\ \mathrm{d}x
\end{aligned}
$$ This integral is a bit tricky, since it includes an absolute value.
We can split it up into two cases, the case where $x$ is negative, and
the case where it isn't:\
$$
\begin{aligned}
\int|x|\ \mathrm{d}x&=\begin{cases}
\int x\ \mathrm{d}x & \text{if }x\geq0 \\
\int -x\ \mathrm{d}x & \text{if }x<0
\end{cases} \\
&=\begin{cases}
\frac{x^2}{2}+C & \text{if }x\geq0 \\
\frac{-x^2}{2}+C & \text{if }x<0
\end{cases} \\
&=\begin{cases}
\frac{x\cdot|x|}{2}+C & \text{if }x\geq0 \\
\frac{x\cdot|x|}{2}+C & \text{if }x<0
\end{cases} \\
&=\frac{x|x|}{2}+C
\end{aligned}
$$ We insert this into the previous integral. $$
\begin{aligned}
\frac{1}{2\theta}\int_{-\theta}^\theta |x|\ \mathrm{d}x&=\frac{1}{2\theta}\left[\frac{x|x|}{2}\right]_{x=-\theta}^{x=\theta} \\
&=\frac{1}{2\theta}\left(\left(\frac{\theta|\theta|}{2}\right)-\left(\frac{-\theta|-\theta|}{2}\right)\right) \\
\end{aligned}
$$ $\theta$ may be negative, so I can't remove the absolute function,
but I can remove the negative sign from the $|-\theta|$, as it will end
up positive no matter what.\
$$
\begin{aligned}
\mathrm{E}[|X_i|]&=\frac{1}{2\theta}\left(\left(\frac{\theta|\theta|}{2}\right)-\left(\frac{-\theta|\theta|}{2}\right)\right) \\
&=\frac{1}{2\theta}\left(\frac{\theta|\theta|}{2}+\frac{\theta|\theta|}{2}\right) \\
&=\frac{1}{2\theta}\left(\frac{2\theta|\theta|}{2}\right) \\
&=\frac{\theta|\theta|}{2\theta} \\
&=\frac{|\theta|}{2} \\
\end{aligned}
$$ Now that we know the expectation of $X_i$, we can insert it into the
expectation for $\hat{\Theta}$ function. $$
\begin{aligned}
\mathrm{E}[\hat{\Theta}]&=\frac{2}{n}(\mathrm{E}[|X_1|]+\mathrm{E}[|X_2|]+\dots+\mathrm{E}[|X_n|]) \\
&=\frac{2}{n}\left(\frac{|\theta|}{2}+\frac{|\theta|}{2}+\dots+\frac{|\theta|}{2}\right) \\
&=\frac{2}{n}\cdot\frac{|\theta|}{2}+\frac{2}{n}\cdot\frac{|\theta|}{2}+\dots+\frac{2}{n}\cdot\frac{|\theta|}{2}) \\
&=\frac{2|\theta|}{2n}+\frac{2|\theta|}{2n}+\dots+\frac{2|\theta|}{2n} \\
&=\frac{|\theta|}{n}+\frac{|\theta|}{n}+\dots+\frac{|\theta|}{n} \\
&=n\cdot\frac{|\theta|}{n} \\
&=|\theta|
\end{aligned}
$$ Now technically we don't know if it actually gives $\theta$, because
if $\theta$ is negative, it gives $-\theta$. This however, implies that
$-\theta>\theta$, which is not allowed for the uniform distribution, as
the lower bound *has* to be lower than the upper bound, meaning $\theta$
cannot be negative. Thus we know that $\theta>-\theta$, and
$\mathrm{E}[\hat{\Theta}]=\theta$.

b.  Now we ant to prove that\
    $$
    T=\frac{3}{n}(X^2_1+X^2_2+\dots+X^2_n)
    $$ is an unbiased estimator for $\theta^2$.\
    $$
    \begin{aligned}
    \mathrm{E}[T]&=\mathrm{E}[\frac{3}{n}(X^2_1+X^2_2+\dots+X^2_n)] \\
    \end{aligned}
    $$ Use linearity of expectations to write $$
    \begin{aligned}
    \mathrm{E}[T]&=\mathrm{E}[\frac{3}{n}(X^2_1+X^2_2+\dots+X^2_n)] \\
    &=\frac{3}{n}(\mathrm{E}[X^2_1]+\mathrm{E}[X^2_2]+\dots+\mathrm{E}[X^2_n]) \\
    \end{aligned}
    $$\
    Now we find the expectation of $X^2_i$, again using the change of
    variable formula.\
    $$
    \begin{aligned}
    \mathrm{E}[X^2_i]&=\int_{-\theta}^\theta x^2\frac{1}{2\theta}\ \mathrm{d}x \\
    &=\frac{1}{2\theta}\int_{-\theta}^\theta x^2\ \mathrm{d}x \\
    &=\frac{1}{2\theta}\left[\frac{x^3}{3}\right]_{-\theta}^\theta \\
    &=\frac{1}{2\theta}\left(\frac{\theta^3}{3}+\frac{\theta^3}{3}\right) \\
    &=\frac{1}{2\theta}\cdot\frac{2\theta^3}{3} \\
    &=\frac{\theta^2}{3}
    \end{aligned}
    $$ Inserting in the previous equation... $$
    \begin{aligned}
    \mathrm{E}[T]&=\frac{3}{n}(\mathrm{E}[X^2_1]+\mathrm{E}[X^2_2]+\dots+\mathrm{E}[X^2_n]) \\
    &=\frac{3}{n}(\frac{\theta^2}{3}+\frac{\theta^2}{3}+\dots+\frac{\theta^2}{3}) \\
    &=\frac{3}{n}\cdot\frac{n\theta^2}{3} \\
    &=\theta^2
    \end{aligned}
    $$ So it is unbiased.

c.  $\sqrt{T}$ is biased, as unbiasedness does not carry over in the
    square root operation, but I'm too tired to prove it right now.\
    I would assume it has a negative bias.

# 2. Maximum likelihood estimator for geometric random variables (Theory)

The geometric random variable, as presented in the textbook, has the
following probability mass function

$$
\mathrm{Pr}[X = k] = (1-p)^{k-1}\cdot p
$$

which can be described as the probability of requiring k trials to
obtain the first success in a sequence of Bernoulli trials. For this
random variable, we have seen that the maximum likelihood estimator for
the parameter p is

$$
\hat{p}=\frac{n}{\sum_{i=0}^n x_i}=\frac{1}{x}
$$

However, in some contexts[^1] a slightly different definition of
geometric random variable is used:

$$
\mathrm{Pr}[X=k]=(1-p)^k\cdot p
$$

This second formulation can be described as the probability of
experiencing k consecutive failures before the first success.

We shall see, with this exercise, that this small change leads to a
different maximum likelihood estimator for $p!$

a. Derive the loglikelihood function $\ell(p)$.

b. Compute the derivative $\ell'(p)$ of the loglikelihood function.

c. Show that the maximum likelihood estimator for $p$ is

$$
\hat{p}=\frac{n}{n+\sum_{i=0}^n x_i}
$$

Therefore, *pay attention* to the distribution you are dealing with,
always read carefully the definitions and the documentation.
  
*Solution*:  
a. The distribution is defined by
$$
\mathrm{P}(X=k)=(1-p)^k\cdot p=f(x)
$$
The likelihood function of the estimator $p$ is then  
$$
\begin{aligned}
L(p)&=p_p(x_1)p_p(x_2)\dots p_p(x_n) \\
&=((1-p)^{x_1}\cdot p)((1-p)^{x_2}\cdot p)\dots((1-p)^{x_n}\cdot p) \\
&=(1-p)^{x_1}p(1-p)^{x_2} p\dots(1-p)^{x_n}p \\
&=p^n(1-p)^{x_1}(1-p)^{x_2}\dots(1-p)^{x_n} \\
&=p^n(1-p)^{x_1+x_2+\dots+x_n}
\end{aligned}
$$
The loglikelihood is just the log-transformed likelihood:
$$
\begin{aligned}
\ell(p)&=\ln L(p) \\
&=\ln \left(p^n(1-p)^{x_1+x_2+\dots+x_n}\right) \\
&=n\ln (p)+(x_1+x_2+\dots+x_n)\ln (1-p) \\
&=n\ln (p)+\left(\sum_{i=0}^n x_i\right)\ln (1-p)
\end{aligned}
$$
  
b. Compute derivative:  
$$
\begin{aligned}
\frac{\mathrm{d}}{\mathrm{d}p}[\ell(p)]&=\frac{\mathrm{d}}{\mathrm{d}p}\left[n\ln (p)+\left(\sum_{i=0}^n x_i\right)\ln (1-p)\right] \\
&=n\frac{\mathrm{d}}{\mathrm{d}p}[\ln(p)]+\frac{\mathrm{d}}{\mathrm{d}p}\left[\left(\sum_{i=0}^n x_i\right)\ln (1-p)\right] \\
&=\frac{n}{p}+\left(\sum_{i=0}^n x_i\right)\frac{1}{p-1}
\end{aligned}
$$
  
c. To find the estimator, we set the above derivative to equal zero and isolate $p$:
$$
\frac{n}{p}+\left(\sum_{i=0}^n x_i\right)\frac{1}{p-1}=0 \Leftrightarrow \\
p=\frac{n}{n+\sum_{i=0}^n x_i}
$$

# 3. Maximum likelihood estimator for geometric random variables (R)

In Problem 2, you showed that the geometric distribution defined as

$$
\mathrm{Pr}[X=k]=(1-p)^k\cdot p
$$

has the following maximum likelihood estimator for p:

$$
\hat{p}^*=\frac{n}{n+\sum_{i=0}^nx_i}.
$$

This definition of geometric random variable is the one use by R, as
stated at the beginning of the "Details" section of `help(rgeom)`. In
this exercise you will verify that, in this case, using the inverse of
the sample mean as the estimator for p leads to heavily biased
estimations.

Let $n = 200$. First of all, define a function `estimate_p` that, given
the realization of a random sample of $n$ elements it returns the
estimate of $p$ using the estimator
$\hat{p}^*=\frac{n}{n+\sum_{i=0}^nx_i}$.

Then, define $p = 0.3$, and take a random sample of $n$ elements using the `rgeom` function. From this random sample, estimate $p$ using first the estimator $\hat{p}=\frac{1}{x}$ and then using the estimator $\hat{p}^*=\frac{n}{n+\sum_{i=0}^n x_i}$. Compute the two values $\hat{p}-p$ and $\hat{p}^*-p$. What do the resulting numbers suggest?

Repeat the above sampling and estimation procedure 1000 times, accumulating the values $\hat{p}-p$ and $\hat{p}^*-p$ in two separate lists. Plot the two distributions, possibly overlaying them on the same plot. What can you conclude by observing the plot?
  
*Solution*:  
```{r}
n <- 200
estimate_p <- function(X) {
  return(length(X)/(length(X)+sum(X)))
}
p <- 0.3

diffs <- c()
diffs_star <- c()
for (i in 1:1000) {
  X <- rgeom(n, p)
  p_est <- 1/mean(X)
  p_est_star <- estimate_p(X)
  diff <- p_est - p
  diff_star <- p_est_star - p
  
  diffs <- c(diffs, diff)
  diffs_star <- c(diffs_star, diff_star)
}
plot(density(diffs_star), xlim=c(-0.06, 0.3), col="magenta", sub="Magenta: True estimate. Blue: Bad estimate")
lines(density(diffs), col="blue")
```
As we can see the true estimate is centered around 0, meaning its expected value is that of the true $p$, while the bad estimate is positively biased and also has a wider standard deviation.  
  
# 4. Linear regression model and residuals (R)  
Let us take a look at the `Cars93 (MASS)` data set.  
(a) Plot the mileage `MPG.highway` in the function of   `Horsepower`. Compute the least-squares estimate for the regression line and add it to the plot.  
(b) What the predicted mileage for a car with 225 horsepower?  
(c) Compute and plot the residuals in the function of horsepower. On the basis of the residuals, is the linear model assumption reasonable?  
  
*Solution*:  
a.  
```{r}
library(MASS)
fit <- lm(Cars93$MPG.highway ~ Cars93$Horsepower)
plot(Cars93$Horsepower, Cars93$MPG.highway)
abline(fit, col="red")
```
b.  
```{r}
coeffs <- coefficients(fit)
estimate_mpg <- function(x) {
  return(coeffs[1] + coeffs[2]*x)
}
hp <- 225
cat("The estimated mileage for a car with", hp, "horsepower is", estimate_mpg(hp))
```
c.  
```{r}
res <- resid(fit)
plot(res)
```
It is homoscedastic and looks evenly distributed.
  
# 5. Linear models (Theory)  
In some situations we may know that the linear model should have some peculiarities, like having no slope, or having intercept equals to zero[^2]. Answer to the two following separate questions (i.e. the answer to one doesn’t depend on the answer to the other). Let $U_i$ be random variables with expectation zero and variance $\sigma^2$.  
  
a. Consider the case $\alpha=0$. The model then becomes $Y_i=\beta x_i+U_i$, for $i=1,2,\dots,n$. Find the least squares estimate $\hat{\beta}$ for $\beta$.  
b. Consider the case $\beta=0$. The mode is then $Y_i=\alpha+U_i$, for $i=1,2,\dots,n$. Find the least squares estimate $\hat{\alpha}$ for $\alpha$.  
  
*Solution*:  
a.  
The formulas for both estimators are:
$$
\begin{aligned}
n\alpha+\beta\sum_{i=1}^nx_i&=\sum_{i=1}^ny_i \\
\alpha\sum_{i=1}^nx_i+\beta\sum_{i=1}^nx_i^2&=\sum_{i=1}^nx_iy_i.
\end{aligned}
$$
When $\alpha=0$, $\hat{\beta}$ is
$$
\begin{aligned}
\beta\sum_{i=1}^nx_i&=\sum_{i=1}^ny_i \Leftrightarrow\\  
\hat{\beta}&=\frac{\sum_{i=1}^n y_i}{\sum_{i=1}^n x_i}.
\end{aligned}
$$
When $\beta=0$, $\hat{\alpha}$ is
$$
\begin{aligned}
\alpha\sum_{i=1}^nx_i&=\sum_{i=1}^nx_iy_i \Leftrightarrow \\
\hat{\alpha}&=\frac{\sum_{i=1}^n x_iy_i}{\sum_{i=1}^n x_i}.
\end{aligned}
$$

[^1]: Including the R implementation of the geometric random distribution.  
[^2]: For instance we may know that when one quantity of the bivariate dataset is $0$ then the other must be zero.