---
title: "APSTA Week 11 Exercises"
output: pdf_document
---
# 1. Sugar Packing (Theory)  
A sugar packaging machine is filling the bags with sugar where the weight accurately follows normal
distribution with the mean $\mu$ and variance $\sigma^2$, where $\sigma=7\text{g}$. In the sample of $36$ packages the sample mean
for the weight is $507\text{g}$. Compute the $97\%$ confidence interval for $\mu$.  
  
*Solution*:  
The formula for the $100(1-\alpha)\%$ confidence interval for normal distributions with known variance is:
$$
\left(\bar{x}_n-z_{\alpha/2}\frac{\sigma}{\sqrt{n}},\bar{x}_n+z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right).
$$
Since we want $93\%$, we set $\alpha=0.03$. The mean is $\bar{x}_n=507$, samples $n=36$ and the standard deviation $\sigma=7$. That makes our $97\%$ confidence interval
$$
\left(507-z_{0.015}\frac{7}{6},507+z_{0.015}\frac{7}{6}\right).
$$
Now we just need to find the critical value $z_{0.015}$. It can be calculated by taking the inverse CDF of the normal distribution at $1-0.015$:
$$
F(z_p)=1-p \Leftrightarrow F^{-1}(1-p)=z_p
$$
This function exists in R as `qnorm`:
```{r}
qnorm(1-0.015)
```
We therefore insert $2.17$ in the formula.
$$
\begin{aligned}
\left(507-2.17\frac{7}{6},507+2.17\frac{7}{6}\right)&=\left(507-\frac{15.19}{6},507+\frac{15.19}{6}\right) \\
&=(504.47,509.53)
\end{aligned}
$$
We are $97\%$ confident that the this interval contains the true mean.  
  
# 2. Bags of potatoes (Theory)  
You bought $10$ very large bags of potatoes. Assume that the $10$ weights can be viewed as a realization of a
random sample from a normal distribution with unknown parameters. Your measures give you the following
data:  
- Sample mean: $14.5\ \text{kg}$  
- Sample standard deviation: $0.3\ \text{kg}$  

Construct a $95\%$ confidence interval for the expected weight of a bag.  
  
*Solution*:  
The formula for the confidence interval for a normal distribution with unknown variance is
$$
\left(\bar{x}_n-t_{n-1,\alpha/2}\frac{s_n}{\sqrt{n}}, \bar{x}_n+t_{n-1,\alpha/2}\frac{s_n}{\sqrt{n}}\right).
$$  
This time, we need to find the critical value of the t-distribution with $n-1$ degrees of freedom. We do the same as above, but this time with the `qt` command. We have $10-1=9$ degrees of freedom, and since we want the $95\%$ confidence interval, our $\alpha=0.05$.
```{r}
qt(0.95, 9)
```
Therefore, our critical value is $1.83$. We insert the values into the interval:  
$$
\begin{aligned}
\left(14.5-1.83\frac{0.3}{\sqrt{10}}, 14.5+1.83\frac{0.3}{\sqrt{10}}\right)=(14.326, 14.674)
\end{aligned}
$$
  
# 3. How many samples do we need? (Theory)  
Assume that we measure a person’s height (in meters) and that the measurements are normal distributed
with standard deviation $\sigma=0.01$. How many measurements do we to make, if we want a $99\%$ confidence
interval no wider than $0.001$ meters for the mean $\mu$? Please explain how you find the number of required
measurements.  
  
*Solution*:  
We can determine the size of the confidence interval from a confidence level based on the formula for the confidence interval. The size for an interval with confidence level $1-\alpha$ is
$$
2\cdot z_{\alpha/2}\frac{\sigma}{\sqrt{n}}.
$$
If we then impose a restriction of a maximum width, we can isolate the number of samples required:
$$
2\cdot z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\leq w\ \Leftrightarrow\ n\geq\left(\frac{2z_{\alpha/2}\sigma}{w}\right)^2.
$$
The maximum width $w$ is $0.001$, and the standard deviation is $\sigma=0.01$. We want a $99\%$ confidence interval, so $\alpha=0.01$. Finally, we need to find the critical value.
```{r}
qnorm(1-(0.01/2))
```
The critical value is 2.576. We can then insert the numbers:
$$
n\geq\left(\frac{2\cdot2.576\cdot0.01}{0.001}\right)^2=51.52
$$
The minimum number of samples is 52.  
  
# 4. Cats & confidence (R)  
The `cats` data set (available in `UsingR`) contains the bodyweight and heart weight of adult cats, along with
their sex.
(a) Compute the mean and the $90\%$ confidence intervals for the body weight and heart weight assuming
normality of the samples. Do the computation separately for the female and male cats. (So you have to
compute $4$ confidence intervals). Is there any difference between the results?  
(b) Compute the one-sided $95\%$ confidence intervals for the mean body weight of male cats and compare to
the results you obtained at (a).  
  
*Solution*:  
I start by creating a function for computing the confidence interval.
```{r}
CI <- function(mean, crit_val, sd, samples) {
  return(
    c(
      mean - crit_val * (sd / sqrt(samples)),
      mean + crit_val * (sd / sqrt(samples))
    )
  )
}
```
Since the true variance is unknown, I need to use the t-distribution. Since I want to $90\%$ confidence interval, $\alpha=0.1$ and $p=\alpha/2=0.05$.  
```{r}
library(UsingR)
p <- 0.05
crit_val <- qnorm(1-p)

cats_F <- cats[cats$Sex == "F",]
cats_M <- cats[cats$Sex == "M",]
sample_means <- c(
  mean(cats_F$Bwt), mean(cats_F$Hwt),
  mean(cats_M$Bwt), mean(cats_M$Hwt)
)
names(sample_means) <- c(
  "Female body mean", "Female heart mean",
  "Male body mean", "Male heart mean"
)
```
Now that I've calculated the means, I can now calculate the confidence intervals.
```{r}
sample_sds <- c(
  sd(cats_F$Bwt), sd(cats_F$Hwt),
  sd(cats_M$Bwt), sd(cats_M$Hwt)
)
names(sample_sds) <- c(
  "Female body sd", "Female heart sd",
  "Male body sd", "Male heart sd"
)

FB_ci <- CI(sample_means[1], crit_val, sample_sds[1], length(cats_F))
FH_ci <- CI(sample_means[2], crit_val, sample_sds[2], length(cats_F))
MB_ci <- CI(sample_means[3], crit_val, sample_sds[3], length(cats_M))
MH_ci <- CI(sample_means[4], crit_val, sample_sds[4], length(cats_M))

cat("90% confidence intervals:")
cat("\nFemale body weight:", FB_ci)
cat("\nFemale heart weight:", FH_ci)
cat("\nMale body weight:", MB_ci)
cat("\nMale heart weight:", MH_ci)
```
The intervals seem to be centered on larger values for males than females. Whatever the "heart weight" is, it's also way larger than the body weight.  
Now I need the $95\%$ confidence interval for for the male body weight. This means $\alpha=0.05$ and $p=\alpha/2=0.025$.
```{r}
p <- 0.025
crit_val <- qnorm(1-p)

ci <- CI(sample_means[3], crit_val, sample_sds[3], length(cats_M))
cat("Male body weight 95%:", ci)
```
As expected, the confidence interval has grown slightly to accommodate our increase confidence.  
  
# 5. Bootstrapping confidence intervals (R)  
Consider again the `cats` dataset of the previous exercise. Construct the $90\%$ confidence intervals for the
mean body weight of male cats by empirical bootstrap, using $500$ bootstrap repetitions. Compare the result
to those you got in Problem 1.  
How do they compare with the intervals you found in the previous exercise?  
  
*Solution*:  
To bootstrap, we first need to create an ECDF, then create studentised bootstrap means.
```{r}
Fn <- ecdf(cats_M$Bwt)
sample_mean <- sample_means[3]
bootstrap_studentised_means <- c()
for (i in 1:500) {
  bootstrap_sample <- sample(cats_M$Bwt, size = length(cats_M), replace = TRUE)
  bootstrap_mean <- mean(bootstrap_sample)
  bootstrap_sd <- sd(bootstrap_sample)
  
  bootstrap_studentised_means <- c(
    bootstrap_studentised_means,
    ( bootstrap_mean - sample_mean ) / ( bootstrap_sd / sqrt(length(cats_M)) )
  )
}
```
Now that we have our studentised bootstrap means, we find $c_l^*$ and $c_u^*$ such that
$$
\mathrm{P}\left(c_l^*<\frac{\bar{X}_n^*-\mu^*}{S_n^*/\sqrt{n}}<c_u^*\right)\approx1-\alpha.
$$
I do so using the inverse ECDF. This is simply the quantile function. Since I want a $90\%$ confidence interval, I use the $0.05th$ and $0.95th$ quantiles:
```{r}
cl <- quantile(bootstrap_studentised_means, .05)
cu <- quantile(bootstrap_studentised_means, .95)
```
These values can then be inserted into the following formula:
$$
\left(\bar{x}_n-c_u^*\frac{s_n}{\sqrt{n}},\bar{x}_n-c_l^*\frac{s_n}{\sqrt{n}}\right).
$$
```{r}
interval <- c(
  sample_mean - cu * ( sample_sds[3] / sqrt(length(cats_M)) ),
  sample_mean - cl * ( sample_sds[3] / sqrt(length(cats_M)) )
)
cat("The 90% bootstrapped confidence interval for male body weight is", interval)
```
For easier viewing, here are the three intervals:
$$
\begin{aligned}
\text{Male body weight, 90\%: } &(2.456,3.344) \\
\text{Male body weight, 95\%: } &(2.371, 3.429) \\
\text{Male body weight, 90\% bootstrapped: } &(2.124, 3.841)
\end{aligned}
$$
As we can see, the bootstrapped interval is significantly wider than the other two, even when comparing it to the $95\%$ confidence interval. This just goes to show how much better your confidence becomes when you *know* what distribution your data is coming from.