---
title: "Various exam questions"
output: pdf_document
---
2 Football Betting (8pt)
Imagine a football betting setting where there are 13 football games. Each game can have three possible
outcomes: home team wins (1), the teams play even (E) or the visitor team wins (2). Model the outcome
of each game as a random process where each of the three outcomes are equally probable and independent
from other games. Let the random variable X characterise the number of correct guesses for the 13
outcomes in one betting.
a) (Ipt) Write the analytic forms for the probability mass function of X.  
*Solution*:  
bla bla
$$
p(a)=\left(\frac{2}{3}\right)^{13-a}\cdot\left(\frac{1}{3}\right)^a
$$
  
b) (Ipt) Illustrate the probability mass function by plotting it in a figure.  
*Solution*:  
```{r}
prob <- function(a) {
  return(
    (
      (2/3)^(13-a)
    ) * (
      (1/3)^a
    )
  )
}

probs <- c()
for (i in 0:13) {
  prub <- prob(i)
  probs <- c(probs, prub)
}

plot(y=probs, x=0:13)
```
  

c) (Ipt) What is the probability that one get all the 13 outcomes right?  
*Solution*:  
```{r}
prob(13) 
```
  
d) (3pt) Simulate the betting by "playing" the betting 100 times. Present the results you got.  
*Solution*:  
```{r}
game_result <- function() {
  res <- runif(1, min=-1, max=1)
  return(round(res))
}

setting_result <- function() {
  game_results <- c()
  for (i in 1:13) {
    game_results <- c(game_results, game_result())
  }
  return(game_results)
}

bet_res <- c()
for (i in 1:100) {
  set <- setting_result()
  bet <- setting_result()
  bet_res <- c(bet_res, sum(set == bet))
}

summary(bet_res)
sd(bet_res)
plot(density(bet_res))
```
  
e) (2pt) Assume, one betting costs 0.5 EUR. How would you characterise your chances of getting profit by betting in these games by guessing the results in random? Assume that wins are distributed
according to the table below (price is in EUR).  
```
| Hits | Price  |
| 13   | 113101 | 
| 12   | 7761   |
| 11   | 373    |
| 10   | 63     |
```
*Solution*:  
```{r}
total_cost <- 100*0.5
total_profit <- 0
for (i in 1:100) {
  res <- bet_res[i]
  if (res == 10) {
    total_profit = total_profit + 63
  } else if (res == 11) {
    total_profit = total_profit + 373
  } else if (res == 12) {
    total_profit = total_profit + 7761
  } else if (res == 13) {
    total_profit = total_profit + 113101
  }
}
net_gain <- total_profit - total_cost
sprintf("Total profit: %s", total_profit)
sprintf("Net gain: %s", net_gain)
```
Expected value calculation. First make new random variable, $Y$. It a discrete random variable with same distribution as $X$ that gives the money gained from a betting round. It takes the numbers 0 to 13, indicating number of correct guesses, and gives out the prizes defined in the table above.
$$
\begin{aligned}
E[Y]&=\sum_i a_ip(a) \\
&=\left(\frac{2}{3}\right)^{13}\cdot\left(\frac{1}{3}\right)^0\cdot0+\left(\frac{2}{3}\right)^{12}\cdot\left(\frac{1}{3}\right)^1\cdot0+\dots \\
&+\left(\frac{2}{3}\right)^{1}\cdot\left(\frac{1}{3}\right)^{12}\cdot7761+\left(\frac{2}{3}\right)^0\cdot\left(\frac{1}{3}\right)^{13}\cdot113101 \\
&=\left(\frac{2}{3}\right)^3\cdot\left(\frac{1}{3}\right)^{10}\cdot63+\left(\frac{2}{3}\right)^2\cdot\left(\frac{1}{3}\right)^{11}\cdot373 \\
&+\left(\frac{2}{3}\right)^1\cdot\left(\frac{1}{3}\right)^{12}\cdot7761+\left(\frac{2}{3}\right)^0\cdot\left(\frac{1}{3}\right)^{13}\cdot113101 \\
&=\frac{130619}{1594323}\approx0.0819
\end{aligned}
$$
Expected value for playing is 0.0819 euro. Seeing as it also costs 0.5 euro to play, it means that our actual expected value is $0.0819-0.5=-0.4181$, that is, we loose money. Don't gamble kids.  
  
# Maximum likelihood
Let $x_1, x_2, x_3, \dots, x_n$ be a dataset where all the observations are independent and identically distributed
(i.i.d.) following a continuous distribution with the probability density function $f_a(x)$ given by  
$$
f_a(x)=
\begin{cases}
  \mathrm{e}^{-(x-\alpha)}&\text{for }x\geq\alpha, \\
  0, & x<\alpha,
\end{cases}
$$
where $\alpha$ is a parameter.  
(a) (2pt) Write down the likelihood function $L(\alpha)$ corresponding to the dataset above.  
*Solution*:  
The likelihood function is given by
$$
\begin{aligned}
  f_\alpha(x_1)f_\alpha(x_2)f_\alpha(x_3)\dots f_\alpha(x_n)= \\
  \begin{cases}
    \mathrm{e}^{-(x_1-\alpha)}\mathrm{e}^{-(x_2-\alpha)}\dots\mathrm{e}^{-(x_n-\alpha)} & \text{if } x_i\geq\alpha~\forall~x, \\
    0 & \text{if } x_i<\alpha~\forall~x
  \end{cases}
\end{aligned}
$$
  
(b) (4pts) Determine the maximum likelihood estimate for $\alpha$.  
*Solution*:
To find the MLE, we need to get the derivative of the likelihood function.  
$$
\begin{aligned}
\frac{dL(\alpha)}{d\alpha}&=\frac{d}{d\alpha}\left[\mathrm{e}^{-x_1+\alpha}\cdot\mathrm{e}^{-x_2+\alpha}\dots\mathrm{e}^{-x_n+\alpha}\right] \\
&=\frac{d}{d\alpha}\left[\mathrm{e}^{-x_1+\alpha}\right]\cdot\frac{d}{d\alpha}\left[\mathrm{e}^{-x_2+\alpha}\right]\dots\frac{d}{d\alpha}\left[\mathrm{e}^{-x_n+\alpha}\right] \\
&=\mathrm{e}^{\alpha-x_1}\mathrm{e}^{\alpha-x_2}\dots\mathrm{e}^{\alpha-x_n}
\end{aligned}
$$
We can take the loglikelihood to simplify the equation:
$$
\begin{aligned}
\ell(\alpha)&=\ln\left(\frac{d}{d\alpha}\left[L(\alpha)\right]\right) \\
&=\ln\left(\mathrm{e}^{\alpha-x_1}\mathrm{e}^{\alpha-x_2}\dots\mathrm{e}^{\alpha-x_n}\right) \\
&=\ln\left(\mathrm{e}^{\alpha-x_1}\right)+\ln\left(\mathrm{e}^{\alpha-x_2}\right)+\dots+\ln\left(\mathrm{e}^{\alpha-x_n}\right) \\
&=\alpha-x_1+\alpha-x_2+\dots+\alpha-x_n \\
&=n\cdot\alpha-x_1-x_2-\dots-x_n
\end{aligned}
$$
Now we set it equal to zero and solve for $\alpha$.  
$$
\begin{aligned}
n\cdot\alpha-x_1-x_2-\dots-x_n&=0 \Leftrightarrow \\
n\cdot\alpha&=x_1+x_2+\dots+x_n \Leftrightarrow \\
\alpha&=\frac{x_1+x_2+\dots+x_n}{n} \Leftrightarrow \\
\alpha&=\bar{x}_n
\end{aligned}
$$
The only critical point in this function is when $\alpha$ equals the sample mean. Since that is the only point, it must be the maximum. 